---
title: "network_comparisons"
author: "Aurelien Dugourd"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

However, the legacy network optimisation of COSMOS was done through CARNIVAL, which requires the interactive version of IBM Cplex or CBC-COIN solver as the network optimizer. Thus, the classic cosmos run using the CARNIVAL back end is still presnet in this tutorial as a comparison reference, but is optional. IF you wish to reproduce it, The IBM ILOG Cplex is freely available through Academic Initiative [here](https://community.ibm.com/community/user/datascience/blogs/xavier-nodet1/2020/07/09/cplex-free-for-students). As an alternative, the CBC solver is open source and freely available for any user, but has a significantly lower performance than CPLEX. The CBC executable can be find under cbc/. Alternatively for small networks, users can rely on the freely available lpSolve R-package, which is automatically installed with the package.

In this tutorial we use *CPLEX* for means of comparison with the newer MOON function that doesn't rely on optimisation.

```{r}
library(readr)
library(dplyr)
library(cosmosR)
```

```{r}
data("HMDB_mapper_vec")

moon_res_rec_to_TFmet <- as.data.frame(
  read_csv("results/cosmos/moon/moon_res_rec_to_TFmet.csv"))
names(moon_res_rec_to_TFmet)[1] <- "source"
moon_res_TF_lig <- as.data.frame(
  read_csv("results/cosmos/moon/moon_res_TF_lig.csv"))

moon_combined <- as.data.frame(rbind(moon_res_rec_to_TFmet, moon_res_TF_lig))
moon_combined <- moon_combined %>% group_by(source) %>% summarise_each(funs(mean(., na.rm = TRUE)))
moon_combined <- as.data.frame(moon_combined)

write_csv(moon_combined, file = "results/cosmos/moon/full_moon_res_combined.csv")

SIF_COSMOS_combined_reduced <- as.data.frame(
  read_csv("results/cosmos/SIF_mofamoon_combined_reduced.csv"))

ATT_COSMOS_combined_reduced <- as.data.frame(
  read_csv("results/cosmos/ATT_mofamoon_combined_reduced.csv"))

meta_network_rec_to_TFmetab <- as.data.frame(
  read_csv("results/cosmos/moon/meta_network_filtered.csv"))
meta_network_TF_lig <- as.data.frame(
  read_csv("results/cosmos/moon/meta_network_TF_lig.csv"))
combined_meta_network <- as.data.frame(
  rbind(meta_network_rec_to_TFmetab,meta_network_TF_lig)
)

combined_meta_network[, 1] <- sapply(combined_meta_network[, 1], function(x, HMDB_mapper_vec) {
        x <- gsub("Metab__", "", x)
        x <- gsub("^Gene", "Enzyme", x)
        suffixe <- stringr::str_extract(x, "_[a-z]$")
        x <- gsub("_[a-z]$", "", x)
        if (x %in% names(HMDB_mapper_vec)) {
            x <- HMDB_mapper_vec[x]
            x <- paste("Metab__", x, sep = "")
        }
        if (!is.na(suffixe)) {
            x <- paste(x, suffixe, sep = "")
        }
        return(x)
    }, HMDB_mapper_vec = HMDB_mapper_vec)

combined_meta_network[, 2] <- sapply(combined_meta_network[, 2], function(x, HMDB_mapper_vec) {
        x <- gsub("Metab__", "", x)
        x <- gsub("^Gene", "Enzyme", x)
        suffixe <- stringr::str_extract(x, "_[a-z]$")
        x <- gsub("_[a-z]$", "", x)
        if (x %in% names(HMDB_mapper_vec)) {
            x <- HMDB_mapper_vec[x]
            x <- paste("Metab__", x, sep = "")
        }
        if (!is.na(suffixe)) {
            x <- paste(x, suffixe, sep = "")
        }
        return(x)
    }, HMDB_mapper_vec = HMDB_mapper_vec)

write_csv(combined_meta_network, file = "results/cosmos/moon/combined_meta_network_translated.csv")
```

```{r}
sum(ATT_COSMOS_combined_reduced$Nodes %in% moon_combined$source)

ATT_COSMOS_combined_reduced[!(ATT_COSMOS_combined_reduced$Nodes %in% moon_combined$source),]
```
Some nodes are not in moon network. They were pruned out by moon because their estimation score as input didn't match the infered activity by moon.

```{r}
compare_moon_carnival <- merge(ATT_COSMOS_combined_reduced[,c(1,9)],moon_combined,by.x = "Nodes",by.y = "source")
compare_moon_carnival$sign_coherence <- sign(compare_moon_carnival$sign_activity) == sign(compare_moon_carnival$score)

sum(compare_moon_carnival$sign_coherence)
sum(compare_moon_carnival$sign_coherence) / dim(compare_moon_carnival)[1]
```

```{r}
write_csv(combined_meta_network, file = "results/cosmos/moon/meta_network_combined_translated.csv")

to_map <-  merge(ATT_COSMOS_combined_reduced[,c(1,9)],moon_combined,by.x = "Nodes",by.y = "source", all.y = T)
names(to_map) <- c("Nodes","CARNIVAL","moon","moon_level")

write_csv(combined_meta_network, file = "results/cosmos/moon/meta_network_combined_translated.csv")
write_csv(to_map, file = "results/cosmos/moon/ATT_to_compare.csv")

```

```{r}
nodes_not_carnival <- unique(c(combined_meta_network$source,combined_meta_network$target))
nodes_not_carnival <- nodes_not_carnival[!nodes_not_carnival %in% ATT_COSMOS_combined_reduced$Nodes]

nodes_predicted_carnival <- ATT_COSMOS_combined_reduced[ATT_COSMOS_combined_reduced$measured == 0,1]

mean(abs(moon_combined[moon_combined$source %in% nodes_predicted_carnival,"score"]))
mean(abs(moon_combined[moon_combined$source %in% nodes_not_carnival,"score"]))
```

#consistent nodes regardless of moon threshold
```{r}
compare_moon_carnival_predicted <- compare_moon_carnival[compare_moon_carnival$Nodes %in% nodes_predicted_carnival,]

sum(compare_moon_carnival_predicted$sign_coherence)
sum(compare_moon_carnival_predicted$sign_coherence) / dim(compare_moon_carnival_predicted)[1]
```

```{r}
combined_SIF_moon <- as.data.frame(read_csv("results/cosmos/moon/combined_SIF_moon.csv"))
names(combined_SIF_moon)[c(1,2,3)] <- c("Node1","Node2","Sign")
combined_ATT_moon <- as.data.frame(read_csv("results/cosmos/moon/combined_ATT_moon.csv"))

length(intersect(combined_ATT_moon$Nodes, ATT_COSMOS_combined_reduced$Nodes))
```

```{r}
SIF_carnival_moon <- as.data.frame(rbind(combined_SIF_moon[,c(1,2,3)], SIF_COSMOS_combined_reduced[,c(1,2,3)]))
# SIF_carnival_moon$edgeID <- paste(SIF_carnival_moon$Node1,SIF_carnival_moon$Node2, sep = "_")
SIF_carnival_moon <- unique(SIF_carnival_moon)

ATT_carnival_moon <- merge(combined_ATT_moon[,c(1,2)],ATT_COSMOS_combined_reduced[,c(1,6)], by = "Nodes", all = T)
names(ATT_carnival_moon)[c(2,3)] <- c("MOON","CARNIVAL")
ATT_carnival_moon$MOON <- sign(ATT_carnival_moon$MOON)
ATT_carnival_moon$CARNIVAL <- sign(ATT_carnival_moon$CARNIVAL)
ATT_carnival_moon[is.na(ATT_carnival_moon$MOON),"MOON"] <- 0
ATT_carnival_moon[is.na(ATT_carnival_moon$CARNIVAL),"CARNIVAL"] <- 0

write_csv(SIF_carnival_moon, file = "results/cosmos/moon/SIF_carnival_moon.csv")
write_csv(ATT_carnival_moon, file = "results/cosmos/moon/ATT_carnival_moon.csv")
```

Strictly consistent between the two network
```{r}
sum(ATT_carnival_moon$CARNIVAL == ATT_carnival_moon$MOON) / length(ATT_carnival_moon[,1])
```

Signficant in carnival network and 0 in the moon
```{r}
sum(ATT_carnival_moon$CARNIVAL != 0 & ATT_carnival_moon$MOON == 0) / length(ATT_carnival_moon[,1])
```
Signficant in moon network and 0 in the carnival
```{r}
sum(ATT_carnival_moon$CARNIVAL == 0 & ATT_carnival_moon$MOON != 0) / length(ATT_carnival_moon[,1])
```

Strickly different between the two networks
```{r}
sum(ATT_carnival_moon$CARNIVAL != 0 & ATT_carnival_moon$MOON != 0 & ATT_carnival_moon$CARNIVAL != ATT_carnival_moon$MOON) / length(ATT_carnival_moon[,1])
sum(ATT_carnival_moon$CARNIVAL != 0 & ATT_carnival_moon$MOON != 0 & ATT_carnival_moon$CARNIVAL != ATT_carnival_moon$MOON) / sum(ATT_carnival_moon$CARNIVAL != 0)
sum(ATT_carnival_moon$CARNIVAL != 0 & ATT_carnival_moon$MOON != 0 & ATT_carnival_moon$CARNIVAL != ATT_carnival_moon$MOON) / sum(ATT_carnival_moon$MOON != 0)
```
% of nodes of carnival that are found in moon and % of nodes of moon that are found in carnival
```{r}
sum(ATT_carnival_moon$CARNIVAL != 0 & ATT_carnival_moon$MOON != 0 & ATT_carnival_moon$CARNIVAL == ATT_carnival_moon$MOON) / sum(ATT_carnival_moon$CARNIVAL != 0)
sum(ATT_carnival_moon$CARNIVAL != 0 & ATT_carnival_moon$MOON != 0 & ATT_carnival_moon$CARNIVAL == ATT_carnival_moon$MOON) / sum(ATT_carnival_moon$MOON != 0)
```

