---
title: "MOFA-COSMOS pipeline"
author: "Pascal Lafrenz and Aurelien Dugourd"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: gfm
---

## MOFA-COSMOS pipeline

```{r Setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Installation and dependency

COSMOS is dependent on CARNIVAL for exhibiting the signalling pathway optimization. CARNIVAL requires the interactive version of IBM Cplex or CBC-COIN solver as the network optimizer. The IBM ILOG Cplex is freely available through Academic Initiative [here](https://community.ibm.com/community/user/datascience/blogs/xavier-nodet1/2020/07/09/cplex-free-for-students). As an alternative, the CBC solver is open source and freely available for any user, but has a significantly lower performance than CPLEX. The CBC executable can be find under cbc/. Alternatively for small networks, users can rely on the freely available lpSolve R-package, which is automatically installed with the package.

In this tutorial we use *CPLEX* which is strongly recommended.

```{r COSMOS installation, message=FALSE}
# Install cosmosR from bioconductor (stable version)
#if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
#BiocManager::install("cosmosR")

# Or install cosmosR from github directly to obtain the newest version
if (!requireNamespace("devtools", quietly = TRUE)) install.packages("devtools")
if (!requireNamespace("cosmosR", quietly = TRUE)) devtools::install_github("saezlab/cosmosR")
```

We are using MOFA2 (Argelaguet et al., 2018) to find correlation between different omics and use its output (factors) to restrict COSMOS input. However, since we are using the python version of MOFA2, please make sure to have [mofapy2](https://github.com/bioFAM/mofapy2) as well as panda and numpy installed in your (conda) environment (e.g. by using reticulate: reticulate::use_condaenv("base", required = T) %\>% reticulate::conda_install(c("mofapy2","panda","numpy")). For downstream analysis, the R version of MOFA2 is used.

```{r MOFA2 installation, message=FALSE}
# Install MOFA2 from bioconductor (stable version)
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("MOFA2", quietly = TRUE)) BiocManager::install("MOFA2")
```

General information (tutorials) on how to use COSMOS and MOFA2, see [COSMOS](https://saezlab.github.io/cosmosR/articles/tutorial.html) and [MOFA2](https://biofam.github.io/MOFA2/tutorials.html).

For data manipulation, visualization, etc. further packages are needed which are loaded here along with MOFA2 and COSMOS.

```{r Loading packages, message=FALSE}
library(cosmosR)
library(MOFA2)
library(readr)
library(ggplot2)
library(ggfortify)
library(dplyr)
library(reshape2)
library(liana)
library(decoupleR)
library(moon)
library(pheatmap)
library(gridExtra)
library(liana)
library(GSEABase)
library(tidyr)
library(RCy3)
```

### From omics data to MOFA ready input

Since extensive pre-processing is beyond the scope of this tutorial, here only the transformation of pre-processed single omics data to a long data.frame (columns: "sample", "group", "feature", "view", "value") is shown. For more details and information regarding appropriate input, please refer to the MOFA tutorial [MOFA2: training a model in R](https://raw.githack.com/bioFAM/MOFA2_tutorials/master/R_tutorials/getting_started_R.html). The raw data is accessed through [NCI60 cellminer](https://discover.nci.nih.gov/cellminer/home.do) and the scripts for pre-processing can be found in the associated folder.

Here, we have the following views (with samples as columns and genes as rows): transcriptomics (RNA-seq [<PMID:31113817>](https://pubmed.ncbi.nlm.nih.gov/31113817/)), proteomics (SWATH (Mass spectrometry) [<PMID:31733513>](https://pubmed.ncbi.nlm.nih.gov/31733513/>)} and metabolomics (LC/MS & GC/MS (Mass spectrometry) [DTP NCI60 data](https://wiki.nci.nih.gov/display/NCIDTPdata/Molecular+Target+Data)). The group information is stored inside the RNA metadata and based on transcription factor clustering (see scripts/RNA/Analyze_RNA.R). Further, only samples are kept for which each omic is available (however this is not necessary since MOFA2 is able to impute data).

```{r Preparing MOFA input, message=FALSE}
# Transcriptomics
## Load data
RNA <- as.data.frame(read_csv("data/RNA/RNA_log2_FPKM_clean.csv"))
rownames(RNA) <- RNA[,1]
RNA <- RNA[,-1]

## Remove genes with excessive amount of NAs (only keep genes with max. amount of NAs = 33.3 % across cell lines)
RNA <- RNA[rowSums(is.na(RNA))<(dim(RNA)[2]/3),]

## Only keep highly variable genes (as suggested by MOFA): here top ≈70% genes to keep >75% of TFs (103 out of 133). For stronger selection, we can further reduce the number of genes (e.g. top 50%)
RNA_sd <- sort(apply(RNA, 1, function(x) sd(x,na.rm = T)), decreasing = T)
hist(RNA_sd, breaks = 100)
hist(RNA_sd[1:6000], breaks = 100)
RNA_sd <- RNA_sd[1:6000]
RNA <- RNA[names(RNA_sd),]

# Proteomics
## Load data
proteo <- as.data.frame(read_csv("data/proteomic/Prot_log10_SWATH_clean.csv"))
rownames(proteo) <- proteo[,1]
proteo <- proteo[,-1]

## Only keep highly variable proteins (as suggested by MOFA): here top ≈60%. For stronger selection, we can further reduce the number of proteins (e.g. only keep top 50%)
proteo_sd <- sort(apply(proteo, 1, function(x) sd(x,na.rm = T)), decreasing = T)
hist(proteo_sd, breaks = 100)
hist(proteo_sd[1:round(dim(proteo)[1]*3/5)], breaks = 100)
proteo_sd <- proteo_sd[1:round(dim(proteo)[1]*3/5)]
proteo <- proteo[names(proteo_sd),]

# Metabolomics
## Load data
metab <- as.data.frame(read_csv("data/metabolomic/metabolomic_clean_vsn.csv"))
rownames(metab) <- metab[,1]
metab <- metab[,-1]

## Since we have only a limited number of metabolite measurements, all measurements are kept here
metab_sd <- apply(metab, 1, function(x) sd(x,na.rm=T))
hist(metab_sd, breaks = 100)

# Create long data frame
## Only keep samples with each view present 
overlap_patients <- intersect(intersect(names(RNA),names(proteo)),names(metab))
RNA <- RNA[,overlap_patients]
proteo <- proteo[,overlap_patients]
metab <- metab[,overlap_patients]

## Create columns required for MOFA
### RNA
RNA <- melt(as.data.frame(cbind(RNA,row.names(RNA))))
RNA$view <- "RNA"
RNA <- RNA[,c(2,1,4,3)]                 
names(RNA) <- c("sample","feature","view","value")

### Proteomics
proteo <- melt(as.data.frame(cbind(proteo,row.names(proteo))))
proteo$view <- "proteo"
proteo <- proteo[,c(2,1,4,3)]                 
names(proteo) <- c("sample","feature","view","value")

### Metabolomics
metab <- melt(as.data.frame(cbind(metab,row.names(metab))))
metab$view <- "metab"
metab <- metab[,c(2,1,4,3)]                 
names(metab) <- c("sample","feature","view","value")

## Merge long data frame to one
mofa_ready_data <- as.data.frame(do.call(rbind,list(RNA,proteo,metab)))

## Add metadata information (cluster assignment)
meta_data <- read_csv("data/metadata/RNA_metadata_cluster.csv")[,c(1,2)]
colnames(meta_data) <- c("sample","cluster")
mofa_ready_data <- merge(mofa_ready_data, meta_data, by = "sample")
mofa_ready_data <- mofa_ready_data[,c(1,5,2,3,4)]
names(mofa_ready_data) <- c("sample","group","feature","view","value")

# Rename clusters
mofa_ready_data[grepl("1",mofa_ready_data$group),2] <- "cluster_1"
mofa_ready_data[grepl("2",mofa_ready_data$group),2] <- "cluster_2"
mofa_ready_data[grepl("3",mofa_ready_data$group),2] <- "cluster_3"

# Optional: Only keep cluster 1 and 3
#mofa_ready_data <- mofa_ready_data[which(mofa_ready_data$group %in% c("1","3")),]

# Here: Remove group column for unsupervised analysis
mofa_ready_data <- mofa_ready_data[,-2]

# Save MOFA metadata information and MOFA input
write_csv(mofa_ready_data, file = "data/mofa/mofa_data.csv")
```

The final data frame has the following structure:

```{r MOFA_input}
head(mofa_ready_data)
```

We have successfully combined single omic data sets to a long multi-omics data frame.

### MOFA run

After pre-processing the data into MOFA appropriate input, let's perform the MOFA analysis. To change specific MOFA options the function write_MOFA_options_file is used. For more information regarding option setting see [MOFA option tutorial](https://raw.githack.com/bioFAM/MOFA2_tutorials/master/R_tutorials/getting_started_R.html). Since we don't know how many factors are needed to explain our data well, various maximum numbers of factors are tested. The final results can be found in the results/mofa/ folder.

```{r MOFA options function, include=FALSE}
write_MOFA_options_file <- function(input_file, 
                                    output_file, 
                                    scale_groups = "False", 
                                    scale_views = "False", 
                                    likelihoods = c('gaussian','gaussian','gaussian','gaussian'),
                                    factors = "5", 
                                    spikeslab_weights = "True",
                                    spikeslab_factors = "False",
                                    ard_factors = "True",
                                    ard_weights = "True",
                                    iter = "1000",
                                    convergence_mode = "fast",
                                    startELBO = "1",
                                    freqELBO = "1",
                                    dropR2 = "0.001", 
                                    gpu_mode = "True", 
                                    verbose = "False", 
                                    seed = "1") {
  wd <- getwd()
  options_MOFA <- data.frame("input_file" = paste0(wd,input_file), 
                        "output_file" = paste0(wd,output_file), 
                        scale_groups, 
                        scale_views,
                        factors,
                        spikeslab_weights,
                        spikeslab_factors,
                        ard_factors,
                        ard_weights,
                        iter,
                        convergence_mode,
                        startELBO,
                        freqELBO,
                        dropR2,
                        gpu_mode,
                        verbose,
                        seed,
                        "length" = length(likelihoods))
  for (i in 1:length(likelihoods)){
    cache <- paste0("likelihood","_",i)
    options_MOFA[,cache] <- likelihoods[i]
  }
  write_csv(options_MOFA, 
            "options_MOFA.csv"
            )
}
```

Do not run this chunk, unless you want to repeat the mofa optimisation, because
it takes a lot of time ot run. The results are already available in the 
results/mofa folder.

```{r MOFA run, eval=FALSE}
for(i in c(5:15,20,length(unique(mofa_ready_data$sample)))){
wd <- getwd()
write_MOFA_options_file(input_file = '/data/mofa/mofa_data.csv', output_file = paste0('/results/mofa/mofa_res_',i,'factor.hdf5'), factors = i, convergence_mode = "slow", likelihoods = c('gaussian','gaussian','gaussian'))
system(paste0(paste('python3', wd, sep = " "), paste('/scripts/mofa/mofa2.py', "options_MOFA.csv")))
}
```

### Investigate MOFA output

We first load the different MOFA models containing the MOFA results together with the metadata information.

```{r Loading MOFA output, message=FALSE, warning=F}
# Load MOFA output
for(i in c(5:15,20,length(unique(mofa_ready_data$sample)))){
  filepath <- paste0('results/mofa/mofa_res_',i,'factor.hdf5')
  model <- load_model(filepath)
  meta_data <- read_csv("data/metadata/RNA_metadata_cluster.csv")[,c(1,2)]
  colnames(meta_data) <- c("sample","cluster")
  mofa_metadata <- merge(samples_metadata(model), meta_data, by = "sample")
  names(mofa_metadata) <- c("sample","group","cluster")
  samples_metadata(model) <- mofa_metadata
  assign(paste0("model",i),model)
}
```

Then, we can compare the factors from different MOFA models and check on consistency (here: model 7-13).

```{r Compare MOFA models}
compare_factors(list(model7,model8,model9,model10,model11,model12,model13), cluster_rows = F, cluster_cols = F,)
```

Here, we can see that the inferred MOFA factor weights do not change drastically around 10. For the sake of this tutorial, we choose the 10-factor MOFA model.

```{r Select MOFA model}
model <- model10
```

The next part deals with the analysis and interpretation of the MOFA factors with specific focus on the factor weights. A more detailed tutorial on how to perform downstream analysis after MOFA model training can be found here [MOFA+: downstream analysis (in R)](https://raw.githack.com/bioFAM/MOFA2_tutorials/master/R_tutorials/downstream_analysis.html). In this context, it is important to emphasize that the factors can be interpreted similarly to principal components (PCs) in a principal component analysis (PCA).

An initial metric we can explore is the total variance ($R^2$) explained per view by our MOFA model. This helps us understand how well the model represents the data.

```{r Total variance per factor}
# Investigate factors: Explained total variance per view
plot_variance_explained(model, x="group", y="factor", plot_total = T)[[2]]
calculate_variance_explained(model)
```

The bar plot demonstrates that the nine factors for the view RNA explain more than 50% of the variance. In contrast, for both metabolomics as well as proteomics around 20% of the variance can be explained by all factors.

Since we would like to use the found correlation for downstream COSMOS analysis, we first investigate the variance ($R^2$) each factor can explain per view as well as investigate the factor explaining the most variance across all views.

```{r Variance per view per factor}
# Investigate factors: Explained variance per view for each factor 
pheatmap(model@cache$variance_explained$r2_per_factor[[1]], display_numbers = T, angle_col = "0", legend_labels = c("0","10", "20", "30", "40", "Variance\n\n"), legend = T, main = "", legend_breaks = c(0,10, 20, 30, 40, max(model@cache$variance_explained$r2_per_factor[[1]])), cluster_rows = F, cluster_cols = F, color = colorRampPalette(c("white","red"))(100), fontsize_number = 10)
pheatmap(model@cache$variance_explained$r2_per_factor[[1]], display_numbers = T, angle_col = "0", legend_labels = c("0","10", "20", "30", "40", "Variance\n\n"), legend = T, main = "", legend_breaks = c(0,10, 20, 30, 40, max(model@cache$variance_explained$r2_per_factor[[1]])), cluster_rows = F, cluster_cols = F, color = colorRampPalette(c("white","red"))(100), fontsize_number = 10,filename = "results/mofa/variance_heatmap.pdf",width = 4, height = 2.5)
```

By analyzing this heatmap, we can observe that Factor 1 and 3 explain a large proportion of variance of the RNA, Factor 2 mainly explains the variance of the metabolomics and Factor 3 highlights variability from the proteomics view. Consequently, no factor explains a large portion of the variance across all views. To potentially justify to use different factor weights downstream for different views, we can analyze the correlation between factors. Thus, the next plot highlights the spearman correlation between each factor.

```{r Correlation between factors}
# Investigate factors: Correlation between factors 
pdf(file="results/mofa/plot_factor_cor.pdf", height = 5, width = 5)
plot_factor_cor(model, type = 'upper', method = "spearman", addCoef.col = "black")
```

Since the correlation between each factor is relatively low, we have to choose a single factor. In this case, we are going to use Factor 4, since a decent amount of variability from each view is explained.

Additionally, we can use a beeswarm plot to examine each sample's factor value and to see if we are able to separate the samples based on their transcription factor clustering.

```{r Factor value per sample and factor}
# Investigate factors: Beeswarm plots of individual factors
plot_factor(model, 
                 factors = 'all',
                 color_by = "cluster",
                 dot_size = 3,
                 dodge = T,           
                 legend = T,          
                 add_violin = T,     
                 violin_alpha = 0.25) + 
   scale_color_manual(values=c("1"="red", "2"="cyan", "3"="orange")) +
   scale_fill_manual(values=c("1"="red", "2"="cyan", "3"="orange"))

plot_factor(model, 
                 factors = 4,
                 color_by = "cluster",
                 dot_size = 3,
                 dodge = T,           
                 legend = T,          
                 add_violin = T,     
                 violin_alpha = 0.25) + 
   scale_color_manual(values=c("1"="red", "2"="cyan", "3"="orange")) +
   scale_fill_manual(values=c("1"="red", "2"="cyan", "3"="orange"))

plot_factor(model, 
                 factors = 'all',
                 dot_size = 3,
                 dodge = T,           
                 legend = T,          
                 add_violin = T,     
                 violin_alpha = 0.25)
```

Interestingly, with Factor 4, we have a separation of the samples belonging to cluster 2 from samples of cluster 1 and 3. Taking into account that this factor mainly explains the variation in the RNA view, the samples are classifiable into the clusters especially by transcriptomics. This is expected since the assignment of the samples was done based on transcription factor clustering via transcriptomics. Thus, it was shown that the original structure of the data was conserved in the low-dimensional representation via factors.

Further, we can inspect the composition of the factors by plotting the top 10 weights per factor and view.

```{r Factor weights per view, fig.height=2.4, fig.width=10}
## Top 20 factor weights per view
grid.arrange(
  ### RNA
  plot_top_weights(model, 
                   view = "RNA",
                   factor = 4,
                   nfeatures = 10) +
    ggtitle("RNA"),
  ### Proteomics
  plot_top_weights(model, 
                   view = "proteo",
                   factor = 4,
                   nfeatures = 10) + 
    ggtitle("Proteomics"),
  ### Metabolomics
  plot_top_weights(model, 
                   view = "metab",
                   factor = 4,
                   nfeatures = 10) + 
    ggtitle("Metabolomics"), ncol =3
)

```

Using this visualization, features with strong association with the factor (large absolute values) can be easily identified and further inspected by literature investigation.

Moreover, we can use heatmaps to highlight the coordinated heterogeneity that MOFA captures in the original data and define clusters by hierarchical clustering. Here, for example, the heatmap for Factor 4 of the RNA view using the top 25 feature is shown and hierarchical clustering with complete linkage is performed. Additionally, the cluster assignment of each sample is depicted.

```{r Heatmap RNA view factor 2}
anno_colors <- list(
  "1"="red", "2"="cyan", "3"="orange"
)

model@samples_metadata$cluster <- as.character(model@samples_metadata$cluster)
### RNA
plot_data_heatmap(model,
                  view = "RNA",       
                  factor = 4,             
                  features =25,          
                  cluster_rows = T, cluster_cols = T,
                  show_rownames = T, show_colnames = F,
                  annotation_samples = "cluster",
                  annotation_colors = anno_colors)
```

Further tools to investigate the latent factors are available inside the MOFA framework ([MOFA+: downstream analysis (in R)](https://raw.githack.com/bioFAM/MOFA2_tutorials/master/R_tutorials/downstream_analysis.html)).

### From MOFA output to TF activity, ligand-receptor activity

In the next step, using the weights of Factor 4 (highest shared $R^2$ across views), different databases (liana and dorothea) and decoupleR, the transcription factor activities, ligand-receptor scores are estimated.

First, we have to extract the weights from our MOFA model and keep the weights from Factor 4 in the RNA view.

```{r Extract weights from Model}
weights <- get_weights(model, views = "all", factors = "all")

# Extract Factor with highest explained variance in RNA view
RNA <- data.frame(weights$RNA[,4]) 
row.names(RNA) <- gsub("_RNA","",row.names(RNA))
```

Then we load the consensus networks from our databases.

```{r Load networks, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
# Load LIANA (receptor and ligand) consensus network
# ligrec_ressource <- distinct(liana::decomplexify(liana::select_resource("Consensus")[[1]]))
# save(ligrec_ressource, file = "support/ligrec_ressource.RData")
load(file = "support/ligrec_ressource.RData")
ligrec_geneset <- ligrec_ressource[,c("source_genesymbol","target_genesymbol")]
ligrec_geneset$set <- paste(ligrec_geneset$source_genesymbol, ligrec_geneset$target_genesymbol, sep = "___")
ligrec_geneset <- reshape2::melt(ligrec_geneset, id.vars = "set")[,c(3,1)]
names(ligrec_geneset)[1] <- "gene"
ligrec_geneset$mor <- 1
ligrec_geneset$likelihood <- 1
ligrec_geneset <- distinct(ligrec_geneset)

# Load Dorothea (TF) network
# dorothea_df <- decoupleR::get_collectri()
# save(dorothea_df, file = "support/dorothea_df.RData")
load(file = "support/dorothea_df.RData")
```

Then, by using decoupleR and prior knowledge networks, we calculate the different regulatory activities inferred by the normalized weighted mean approach. Depending on the task, the minimum number of targets per source varies required to maintain the source in the output (e.g. two targets by definition for ligand-receptor interactions). To calculate activities of master regulons through moon, transcription factor activities estimated by decoupleR and dorothea are used as an input.

```{r RNA across factors, fig.width=4, fig.height=4.3}
source("scripts/support_pheatmap_colors.R")

RNA <- as.data.frame(weights$RNA) 
row.names(RNA) <- gsub("_RNA","",row.names(RNA))

RNA_top <- RNA[order(apply(RNA,1,function(x){max(abs(x))}), decreasing = T)[1:25],]

t <- as.vector(t(RNA_top))
palette1 <- createLinearColors(t[t < 0],withZero = F , maximum = abs(min(t,na.rm = T)) * 10)
palette2 <- createLinearColors(t[t > 0],withZero = F , maximum = abs(max(t,na.rm = T)) * 10)
palette <- c(palette1, palette2)
pheatmap(RNA_top, show_rownames = T, cluster_cols = F, cluster_rows = F,color = palette, angle_col = 315, filename = "results/mofa/mofa_top_RNA.pdf", width = 4, height = 4.3)
pheatmap(RNA_top, show_rownames = T, cluster_cols = F, cluster_rows = F,color = palette, angle_col = 315)
```

```{r LR score estimations across factors, fig.width=4, fig.height=4.3}
# Calculate regulatory activities from Receptor and Ligand network
ligrec_factors <- run_ulm(mat = as.matrix(RNA), network = ligrec_geneset, .source = set, .target = gene, minsize = 2) 
ligrec_factors_df <- reshape2::dcast(ligrec_factors, formula = source~condition, value.var = "score")
row.names(ligrec_factors_df) <- ligrec_factors_df$source
ligrec_factors_df <- ligrec_factors_df[,-1]
ligrec_factors_df <- ligrec_factors_df[order(apply(ligrec_factors_df,1,function(x){max(abs(x))}), decreasing = T)[1:25],]



t <- as.vector(t(ligrec_factors_df))
palette1 <- createLinearColors(t[t < 0],withZero = F , maximum = abs(min(t,na.rm = T)) * 10)
palette2 <- createLinearColors(t[t > 0],withZero = F , maximum = abs(max(t,na.rm = T)) * 10)
palette <- c(palette1, palette2)
pheatmap(ligrec_factors_df, show_rownames = T, cluster_cols = F, cluster_rows = F,color = palette, angle_col = 315, filename = "results/mofa/mofa_top_ligrec.pdf", width = 4, height = 4.3)
pheatmap(ligrec_factors_df, show_rownames = T, cluster_cols = F, cluster_rows = F,color = palette, angle_col = 315)
```

```{r TF activity estimations across factors, fig.width=4, fig.height=4.3}
# Calculate regulatory activities from TF network
TF_factors <- run_ulm(mat = as.matrix(RNA), network = dorothea_df, minsize = 10)
TF_factors <- reshape2::dcast(TF_factors, formula = source~condition, value.var = "score")
row.names(TF_factors) <- TF_factors$source
TF_factors <- TF_factors[,-1]
TF_factors <- TF_factors[order(apply(TF_factors,1,function(x){max(abs(x))}), decreasing = T)[1:25],]

t <- as.vector(t(TF_factors))
palette1 <- createLinearColors(t[t < 0],withZero = F , maximum = abs(min(t,na.rm = T)) * 10)
palette2 <- createLinearColors(t[t > 0],withZero = F , maximum = abs(max(t,na.rm = T)) * 10)
palette <- c(palette1, palette2)
pheatmap(TF_factors, show_rownames = T, cluster_cols = F, cluster_rows = F,color = palette, angle_col = 315, filename = "results/mofa/mofa_top_TF.pdf", width = 4, height = 4.3)
pheatmap(TF_factors, show_rownames = T, cluster_cols = F, cluster_rows = F,color = palette, angle_col = 315)
```


```{r Activity estimations for factor 4}
# Calculate regulatory activities from Receptor and Ligand network
ligrec_high_vs_low <- run_ulm(mat = as.matrix(RNA), network = ligrec_geneset, .source = set, .target = gene, minsize = 2) 
ligrec_high_vs_low <- ligrec_high_vs_low[ligrec_high_vs_low$statistic == "ulm",]
ligrec_high_vs_low_vector <- ligrec_high_vs_low$score
names(ligrec_high_vs_low_vector) <- ligrec_high_vs_low$source

ligrec_high_vs_low_top <- ligrec_high_vs_low[order(abs(ligrec_high_vs_low$score),decreasing = T)[1:15],c(2,4)]
ligrec_high_vs_low_top$source <- factor(ligrec_high_vs_low_top$source, levels = ligrec_high_vs_low_top$source)
ggplot(ligrec_high_vs_low_top, aes(x=source, y = score)) + geom_bar(stat= "identity",position = "dodge") + theme_minimal() + theme(axis.text.x = element_text(angle = 315, vjust = 0.5, hjust=0))


# Calculate regulatory activities from TF network
TF_high_vs_low <- run_ulm(mat = as.matrix(RNA), network = dorothea_df, minsize = 10)
TF_high_vs_low <- TF_high_vs_low[TF_high_vs_low$statistic == "ulm",]
TF_high_vs_low_vector <- TF_high_vs_low$score
names(TF_high_vs_low_vector) <- TF_high_vs_low$source

# Prepare moon analysis input
TF_high_vs_low <- as.data.frame(TF_high_vs_low[,c(2,4)])
row.names(TF_high_vs_low) <- TF_high_vs_low[,1]

TF_high_vs_low_top_10 <- TF_high_vs_low[order(abs(TF_high_vs_low$score), decreasing = T)[1:10],]
TF_high_vs_low_top_10$source <- factor(TF_high_vs_low_top_10$source, levels = TF_high_vs_low_top_10$source)
ggplot(TF_high_vs_low_top_10, aes(x=source, y = score)) + geom_bar(stat= "identity",position = "dodge") + theme_minimal()

# Combine results
ligrec_TF_moon_inputs <- list("ligrec" = ligrec_high_vs_low_vector,
                              "TF" = TF_high_vs_low_vector)

save(ligrec_TF_moon_inputs, file = "data/cosmos/ligrec_TF_moon_inputs.Rdata")
```

We have now successfully used the MOFA weights (Factor 4, RNA view) to infer not only TF activities but also the ligand-receptor interactions.

### Extracting network with COSMOS to formulate mechanistic hypotheses

In this part the COSMOS run is performed. We can use the output of decoupleR (the activities) next to the factor weights as an input for COSMOS to specifically focus the analysis on pre-computed data-driven results.

Here, in the first step, the data is loaded. Besides the activity estimations and the factor weights, the previous filtered out expressed gene names are loaded.

```{r Load COSMOS input, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
# Load data
load(file = "data/cosmos/ligrec_TF_moon_inputs.Rdata")

signaling_input <- ligrec_TF_moon_inputs$TF
ligrec_input <- ligrec_TF_moon_inputs$ligrec

RNA_input <- weights$RNA[,4] 
prot_input <- weights$proteo[,4]
metab_inputs <- weights$metab[,4]

names(RNA_input) <- gsub("_RNA","",names(RNA_input))
names(prot_input) <- gsub("_proteo","",names(prot_input))

expressed_genes <- as.data.frame(read_csv("data/RNA/RNA_log2_FPKM_clean.csv"))$Genes #since only complete cases were considered in the first part, here the full gene list is loaded
```

We can be interested in looking how RNA and corresponding proteins correlate in each factor.
We can plot this using scatter plots for each factors.
```{r correlation RNA/prot, fig.width=7, fig.height=7}
weights_RNA <- as.data.frame(weights$RNA)
weights_prot <- as.data.frame(weights$proteo)

weights_prot$ID <- gsub("_proteo$","",row.names(weights_prot))
weights_RNA$ID <- gsub("_RNA$","",row.names(weights_RNA))

plot_list <- list()
r2_list <- list()
for(i in 1:(dim(weights_prot)[2]-1))
{
  merged_weights <- merge(weights_RNA[,c(i,10)],weights_prot[,c(i,10)], by = "ID")
  names(merged_weights) <- c("ID","RNA","prot")
  r2_list[[i]] <- cor(merged_weights[,2],merged_weights[,3])^2
  plot_list[[i]] <- ggplot(merged_weights,aes(x = RNA,y = prot)) + geom_point() +
  geom_smooth(method='lm', formula= y~x) + 
  theme_minimal() + 
  theme(axis.title.x=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.title.y=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank())
}

r2_vec <- unlist(r2_list)
r2_vec

ggsave(filename = "results/mofa/RNA_prot_cor.pdf",plot = do.call("grid.arrange", c(plot_list, ncol = 3)), device = "pdf")
```
coherently, only the factors where

Next, we perform filtering steps in order to identify top deregulated features. Here, the individual filtering is based on absolute factor weight or activity score. The first step involves the visualization of the respective distribution to define an appropriate threshold.

First, the RNA factor weights are filtered based on a threshold defined by their distribution as well a threshold defined by the distribution of the protein factor weights.

```{r Plot: RNA weights & protein weights}
##RNA
{plot(density(RNA_input))
abline(v = -0.2)
abline(v = 0.2)}

{plot(density(prot_input))
abline(v = -0.05)
abline(v = 0.05)}
```

Based on the plots and indicated by the straight lines, the RNA weights threshold is set to -0.2 and 0.2, and the protein weights threshold to -0.05 and 0.05. RNA factor weight values lying inside this threshold are set to 0 and previously filtered out genes (before MOFA analysis) are also included with value 0.

```{r Filtering RNA weights}
for(gene in names(RNA_input)) {
  if (RNA_input[gene] > -0.2 & RNA_input[gene] < 0.2)
  {
    RNA_input[gene] <- 0
  } else
  {
    RNA_input[gene] <- sign(RNA_input[gene]) * 10
  }
  if (gene %in% names(prot_input))
  {
    if (prot_input[gene] > -0.05 & prot_input[gene] < 0.05)
    {
      RNA_input[gene] <- 0
    } else
    {
      RNA_input[gene] <- sign(RNA_input[gene]) * 10
    }
  }
}

expressed_genes <- expressed_genes[!(expressed_genes %in% names(RNA_input))]
genes <- expressed_genes
expressed_genes <- rep(0,length(expressed_genes))
names(expressed_genes) <- genes

RNA_input <- c(RNA_input, expressed_genes)
```

The same procedure is repeated for the transcription factor activities.

```{r Plot: TF activity estimates}
## Transcription factors
{plot(density(signaling_input))
abline(v = -0.5)
abline(v = 3.5)}
```

The threshold is set to -0.5 and 3.5 respectively. Further, the TF_to_remove variable is later used to remove transcription factors with activities below this threshold from the prior knowledge network.

```{r Filtering TF activity estimates}
TF_to_remove <- signaling_input[signaling_input > -0.5 & signaling_input < 3.5]
signaling_input <- signaling_input[signaling_input < -0.5 | signaling_input > 3.5]
```

The same procedure is repeated for the ligand-receptor activities.

```{r Plot: Ligand-receptor activity estimates}
## Ligand-receptor interactions
{plot(density(ligrec_input))
abline(v = -0.5)
abline(v = 2.5)}
```

Here, only activities higher than 2.5 or lower than -0.5 are kept (see straight lines in plot). Since at this point we are interested in the receptors, the names are adjusted accordingly and if there are multiple mentions of a receptor, its activity is calculated by the mean of the associated ligand-receptor interactions.

```{r Filtering ligand-receptor activity estimates}
ligrec_input <- ligrec_input[ligrec_input > 2.5 | ligrec_input < -0.5]

rec_inputs <- ligrec_input
names(rec_inputs) <- gsub(".+___","",names(rec_inputs))
rec_inputs <- tapply(rec_inputs, names(rec_inputs), mean)
receptors <- names(rec_inputs)
rec_inputs <- as.numeric(rec_inputs)
names(rec_inputs) <- receptors
```


Finally, the same procedure is repeated for the metabolite factor weights.

```{r Plot: Metabolite factor weights}
## Metabolites
{plot(density(metab_inputs))
abline(v = -0.2)
abline(v = 0.2)}
```

The threshold is set to 0.2 and -0.2 respectively. Further, the metab_to_exclude variable is later used to remove metabolites with activities below this threshold from the prior knowledge network. Moreover, metabolite names are translated into HMDB format and metabolic compartment codes (m = mitochondria, c = cytosol) as well as metab\_\_ prefix is added to HMDB IDs. More information regarding compartment codes can be found [here](http://bigg.ucsd.edu/compartments).

```{r Filtering metabolite factor weights, message=FALSE, warning=FALSE, error=FALSE, results='hide'}
metab_to_HMDB <- as.data.frame(
  read_csv("data/metabolomic/MetaboliteToHMDB.csv"))
metab_to_HMDB <- metab_to_HMDB[metab_to_HMDB$common %in% names(metab_inputs),]
metab_inputs <- metab_inputs[metab_to_HMDB$common]
names(metab_inputs) <- metab_to_HMDB$HMDB

metab_inputs <- cosmosR::prepare_metab_inputs(metab_inputs, compartment_codes = c("m","c"))

metab_to_exclude <- metab_inputs[abs(metab_inputs) < 0.2]
metab_inputs <- metab_inputs[abs(metab_inputs) > 0.2]
```

Next, the prior knowledge network (PKN) is loaded. To see how the full meta PKN was assembled, see [PKN](https://github.com/saezlab/meta_PKN_BIGG.git). Since we are not interested in including transcription factors and metabolites that were previously filtered out, we also remove the respective nodes from the PKN.

```{r Meta network, warning=F}
## Load and filter meta network
data("meta_network")

meta_network <- meta_network_cleanup(meta_network)

meta_network <- meta_network[!(meta_network$source %in% names(TF_to_remove)) & !(meta_network$target %in% names(TF_to_remove)),]
meta_network <- meta_network[!(meta_network$source %in% names(metab_to_exclude)) & !(meta_network$target %in% names(metab_to_exclude)),]
```

Then the datasets are merged, entries that are not included in the PKN are removed and the filtering steps are completed. The merging in this case is based on the idea of deriving the forward network via the receptor to transcription factor/metabolite direction.

```{r Merge datasets to input}
upstream_inputs <- c(rec_inputs)
downstream_inputs <- c(metab_inputs*10, signaling_input)

upstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(upstream_inputs, meta_network)
downstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(downstream_inputs, meta_network)

rec_to_TF_cosmos_input <- list(upstream_inputs_filtered, downstream_inputs_filtered)
save(rec_to_TF_cosmos_input, file = "data/cosmos/rec_to_TF_cosmos_input.Rdata")
```

At this point, we can perform the COSMOS analysis. Firstly, the options for the CARNIVAL run such as the time limit and min gap tolerance are set and the user must provide a path to its CPLEX executable. You can check the CARNIVAL_options variable to see all possible options that can be adjusted.

```{r Carnival options pre-run 1, warning=F}
## Set CARNIVAL options to solve pre-optimization problem
my_options <- default_CARNIVAL_options(solver = "cplex")
my_options$solverPath <- "cplex_macos/cplex"
my_options$solver <- "cplex"
my_options$mipGAP <- 0.05
my_options$threads <- 6
my_options$timelimit <- 3200/6
my_options$limitPop <- 100

data("HMDB_mapper_vec")
```

Here, we are trying to find the "forward" network causally connecting the signaling data to metabolic data by using Carnival's ILP solution and the PKN. To further simplify the prior knowledge network (and perform checks on the input data), the pre-processing function of COSMOS is used. Details about the options can be found in the documentation as well as in the [COSMOS tutorial](https://saezlab.github.io/cosmosR/articles/tutorial.html).

```{r COSMOS pre-run forward network 1, eval=FALSE}
pre_run_rec_to_TF <- preprocess_COSMOS_signaling_to_metabolism(meta_network = meta_network,
                                                               signaling_data = upstream_inputs_filtered,
                                                               metabolic_data = downstream_inputs_filtered,
                                                               diff_expression_data = RNA_input,
                                                               maximum_network_depth = 5,
                                                               remove_unexpressed_nodes = T,
                                                               filter_tf_gene_interaction_by_optimization = F,
                                                               CARNIVAL_options = my_options)
```

In this part, we can set up the options for the optimization run. The running time should be much higher here than in pre-optimization. You can increase the number of threads to use if you have many available CPUs.

```{r Carnival options run 1}
## Set CARNIVAL options to solve optimization problem
my_options$mipGAP <- 0.05
my_options$threads <- 6
# my_options$timelimit <- 720
# my_options$limitPop <- 100
```

After pre-optimization, we can perform the actual COSMOS run using the pre-optimized solution.

```{r COSMOS run forward network 1, eval=FALSE}
run_rec_to_TF <- run_COSMOS_signaling_to_metabolism(data = pre_run_rec_to_TF,
                                                    CARNIVAL_options = my_options)
run_rec_to_TF <- cosmosR::format_COSMOS_res(run_rec_to_TF, metab_mapping = HMDB_mapper_vec)
save(run_rec_to_TF, file = "results/cosmos/run_rec_to_TF.Rdata")
```

In order to analyze the gained sub-network, we first process the COSMOS result by extracting the list of interactions (nodes, sign, weight) in the simple interaction format (SIF) and by extracting a list of accompanying attributes (ATT). Here, we can also remove nodes with average activity of 0 as well as with weight of 0.

```{r Processing COSMOS output 1}
## COSMOS output evaluation
load("results/cosmos/run_rec_to_TF.Rdata")

SIF_rec_to_TF <- as.data.frame(run_rec_to_TF[[1]])
SIF_rec_to_TF <- SIF_rec_to_TF[which(SIF_rec_to_TF$Weight != 0),]

ATT_rec_to_TF <- as.data.frame(run_rec_to_TF[[2]])
colnames(ATT_rec_to_TF)[1] <- "Nodes"
ATT_rec_to_TF$measured <- ifelse(ATT_rec_to_TF$NodeType %in% c("M","T","S","P"),1,0)
ATT_rec_to_TF$Activity <- ATT_rec_to_TF$AvgAct
ATT_rec_to_TF <- ATT_rec_to_TF[ATT_rec_to_TF$AvgAct != 0,]

res_mofa_rec_to_TF <- list(SIF_rec_to_TF,ATT_rec_to_TF)
save(res_mofa_rec_to_TF, file = "results/cosmos/res_mofamoon_rec_to_TFmet.RData")
write_csv(SIF_rec_to_TF, file = "results/cosmos/SIF_res_mofamoon_rec_to_TFmet.csv")
write_csv(ATT_rec_to_TF, file = "results/cosmos/ATT_res_mofamoon_rec_to_TFmet.csv")
```

Since we are also interested to derive the forward network via the transcription factor/moon to ligand direction, this COSMOS analysis is additionally performed.

We first assign the activities of the ligand-receptor interactions to the ligands and calculate the mean activity of the ligands in case of multiple entries.

```{r Filtering}
lig_inputs <- ligrec_input
names(lig_inputs) <- gsub("___.+","",names(lig_inputs))
lig_inputs <- tapply(lig_inputs, names(lig_inputs), mean)
ligands <- names(lig_inputs)
lig_inputs <- as.numeric(lig_inputs)
names(lig_inputs) <- ligands
```

Then the datasets are merged appropriately, entries that are not included in the PKN are removed and the filtering steps are completed.

```{r Combine inputs}
upstream_inputs <- c(signaling_input)
downstream_inputs <- c(lig_inputs) 

dorothea_PKN <- dorothea_df[,c(1,3,2)]
names(dorothea_PKN)[2] <- "interaction"

upstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(upstream_inputs, dorothea_PKN)
downstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(downstream_inputs, dorothea_PKN)

TF_to_lig_cosmos_input <- list(upstream_inputs_filtered, downstream_inputs_filtered)
save(TF_to_lig_cosmos_input, file = "data/cosmos/TF_to_lig_cosmos_input.Rdata")
```

```{r Carnival options pre-run 2}
## Set CARNIVAL options to solve optimization problem
my_options$mipGAP <- 0.05
my_options$threads <- 6
my_options$timelimit <- 360
my_options$limitPop <- 100
```

After setting the variables for the CPLEX solver, we can perform the pre-optimization run ...

```{r COSMOS pre-run forward network 2, eval=FALSE}
pre_run_TF_to_lig <- preprocess_COSMOS_signaling_to_metabolism(meta_network = dorothea_PKN, 
                                                               signaling_data = upstream_inputs_filtered,
                                                               metabolic_data = downstream_inputs_filtered,
                                                               diff_expression_data = RNA_input,
                                                               maximum_network_depth = 1,
                                                               remove_unexpressed_nodes = T,
                                                               filter_tf_gene_interaction_by_optimization = T,
                                                               CARNIVAL_options = my_options)
```

... set the options for the actual run ...

```{r Carnival options run 2}
## Set CARNIVAL options to solve optimization problem
my_options$mipGAP <- 0.05
my_options$threads <- 6
my_options$timelimit <- 720
my_options$limitPop <- 100
```

... and perform the optimization.

```{r COSMOS run forward network 2, eval=FALSE}
run_TF_to_lig <- run_COSMOS_signaling_to_metabolism(data = pre_run_TF_to_lig,
                                                    CARNIVAL_options = my_options)

run_TF_to_lig <- cosmosR::format_COSMOS_res(run_TF_to_lig, metab_mapping = HMDB_mapper_vec)
save(run_TF_to_lig, file = "results/cosmos/run_TF_to_lig.Rdata")
```

Again, we extract the information from the determined network and filter out inactive nodes.

```{r Processing COSMOS output 2}
load("results/cosmos/run_TF_to_lig.Rdata")


SIF_TF_to_lig <- as.data.frame(run_TF_to_lig[[1]])
SIF_TF_to_lig <- SIF_TF_to_lig[which(SIF_TF_to_lig$Weight != 0),]


ATT_TF_to_lig <- as.data.frame(run_TF_to_lig[[2]])
colnames(ATT_TF_to_lig)[1] <- "Nodes"
ATT_TF_to_lig$measured <- ifelse(ATT_TF_to_lig$NodeType %in% c("M","T","S","P"),1,0)
ATT_TF_to_lig$Activity <- ATT_TF_to_lig$AvgAct
ATT_TF_to_lig <- ATT_TF_to_lig[ATT_TF_to_lig$AvgAct != 0,]

res_mofa_TF_to_lig <- list(SIF_TF_to_lig,ATT_TF_to_lig)
save(res_mofa_TF_to_lig, file = "results/cosmos/res_mofamoon_TF_to_lig.RData")
write_csv(SIF_TF_to_lig, file = "results/cosmos/SIF_mofamoon_TF_to_lig.csv")
write_csv(ATT_TF_to_lig, file = "results/cosmos/ATT_mofamoon_TF_to_lig.csv")
```

The next step connects both networks, resulting in a complete network of interactions between receptors, ligands, transcription factor regulators, transcription factors, and metabolites.

After merging both inferred networks ATT file, we can identify whether the activity of the node is positive or negative and save the result under the variable sign_activity. Here, to avoid multiple mentions of one specific node, we can combine multiple mentions by calculating the mean.

```{r Combine ATTs, warning=F}
combined_ATT <- as.data.frame(rbind(ATT_rec_to_TF, ATT_TF_to_lig))
combined_ATT$sign_activity <- sign(combined_ATT$AvgAct)

combined_ATT$NodeType <- ifelse(combined_ATT$NodeType == "", 0, 1) #if NodeType is available, set to 1, else set to 0
combined_ATT <- as.data.frame(combined_ATT %>% 
  group_by(Nodes) %>%
  summarise(across(.fns= ~ mean(.x, na.rm = TRUE))))
```

Further, we can also add the MOFA weight and activity score information to the list of attributes.

```{r Add activities and weights to network}
## MOFA weights
MOFA_weights <- get_weights(model, factors = 4, as.data.frame = T)
MOFA_weights <- MOFA_weights %>%
  arrange(feature, view) %>%
  filter(!duplicated(feature))
MOFA_weights <- MOFA_weights[,c(1,3)]
colnames(MOFA_weights) <- c("Nodes", "mofa_weights")

#remove RNA and proteo tags and integrate values
MOFA_weights$Nodes <- gsub("_RNA$","",MOFA_weights$Nodes)
MOFA_weights$Nodes <- gsub("_proteo$","",MOFA_weights$Nodes)
MOFA_weights$sign <- sign(MOFA_weights$mofa_weights)
MOFA_weights$mofa_weights <- abs(MOFA_weights$mofa_weights)

MOFA_weights <- MOFA_weights %>% group_by(Nodes) %>% summarise_each(funs(max(., na.rm = TRUE)))
MOFA_weights <- as.data.frame(MOFA_weights)
MOFA_weights$mofa_weights <- MOFA_weights$mofa_weights * MOFA_weights$sign
MOFA_weights <- MOFA_weights[,-3]

#make metabolite names coherent between mofa and cosmos
#for now i just add the metab input that is already formatted
#I will miss some weights that were not input but will fix that at later time
MOFA_weights_metabaddon <- as.data.frame(metab_inputs)
names(MOFA_weights_metabaddon) <- "mofa_weights"
MOFA_weights_metabaddon$Nodes <- row.names(MOFA_weights_metabaddon)

MOFA_weights_metabaddon[, 2] <- sapply(MOFA_weights_metabaddon[, 2], function(x, HMDB_mapper_vec) {
        x <- gsub("Metab__", "", x)
        suffixe <- stringr::str_extract(x, "_[a-z]$")
        x <- gsub("_[a-z]$", "", x)
        if (x %in% names(HMDB_mapper_vec)) {
            x <- HMDB_mapper_vec[x]
            x <- paste("Metab__", x, sep = "")
        }
        if (!is.na(suffixe)) {
            x <- paste(x, suffixe, sep = "")
        }
        return(x)
    }, HMDB_mapper_vec = HMDB_mapper_vec)

MOFA_weights <- as.data.frame(rbind(MOFA_weights, MOFA_weights_metabaddon))


## Ligands
lig_weights <- ligrec_TF_moon_inputs$ligrec
names(lig_weights) <- gsub("___.+","",names(lig_weights))
lig_weights <- tapply(lig_weights, names(lig_weights), mean)
ligands <- names(lig_weights)
lig_weights <- as.numeric(lig_weights)
names(lig_weights) <- ligands
lig_weights <- data.frame(Nodes = names(lig_weights), feature_weights = lig_weights)

## Receptors
rec_weights <- ligrec_TF_moon_inputs$ligrec
names(rec_weights) <- gsub(".+___","",names(rec_weights))
rec_weights <- tapply(rec_weights, names(rec_weights), mean)
receptors <- names(rec_weights)
rec_weights <- as.numeric(rec_weights)
names(rec_weights) <- receptors
rec_weights <- data.frame(Nodes = names(rec_weights), feature_weights = rec_weights)

## TF
TF_weights <- ligrec_TF_moon_inputs$TF
names(TF_weights) <- gsub("_TF","",names(TF_weights))
TF_weights <- TF_weights[!(names(TF_weights) %in% c(rec_weights$Nodes,lig_weights$Nodes))]
TF_weights <- data.frame(Nodes = names(TF_weights), feature_weights = TF_weights)



## Combine
feature_weights <- as.data.frame(rbind(TF_weights, rec_weights, lig_weights))

## Add weights to data
combined_ATT <- merge(combined_ATT, MOFA_weights, all.x = T)
combined_ATT <- merge(combined_ATT, feature_weights, all.x = T)
```

To identify whether a node is a ligand, receptor or TFr, we first load the consensus PKNs and only keep entries which represent an interaction in our network.

```{r Load consensus}
# LIANA
load("support/ligrec_ressource.RData")
ligrec_df <- ligrec_ressource[,c("source_genesymbol","target_genesymbol")]
ligrec_df <- distinct(ligrec_df)
names(ligrec_df) <- c("Node1","Node2")

ligrec_df$Node1 <- gsub("-","_",ligrec_df$Node1)
ligrec_df$Node2 <- gsub("-","_",ligrec_df$Node2)
ligrec_df$Sign <- 1
ligrec_df$Weight <- 1

ligrec_df <- ligrec_df[ligrec_df$Node1 %in% combined_ATT$Nodes & ligrec_df$Node2 %in% combined_ATT$Nodes, ]

# Load Dorothea (TF) network
load("support/dorothea_df.RData")
dorothea_df <- dorothea_df[dorothea_df$source %in% combined_ATT$Nodes,]
```

Then, we add the information to the network (ligand = 2, receptor = 3, transcription factor = 4, other = 1, no NodeType available = 0).

```{r Set NodeType}
combined_ATT$NodeType <- ifelse(combined_ATT$Nodes %in% ligrec_df$Node1, 2, ifelse(combined_ATT$Nodes %in% ligrec_df$Node2, 3, ifelse(combined_ATT$Nodes %in% dorothea_df$source, 4, combined_ATT$NodeType))) #if node is a ligand set NodeType to 2, if node is a receptor set NodeType to 3, if node is a TF set NodeType to 4, else keep 0 or 1

write_csv(combined_ATT, file = "results/cosmos/ATT_mofamoon_combined.csv")
```

Finally, we merge both SIF together by summarizing multiple entries via mean calculation and add found interactions from the filtered LIANA consensus PKN.

```{r Combine SIFs}
combined_SIF <- as.data.frame(rbind(SIF_rec_to_TF, SIF_TF_to_lig))
combined_SIF <- as.data.frame(combined_SIF %>%
  group_by(Node1, Node2) %>%
  summarise(across(.fns= ~ mean(.x, na.rm = TRUE))))

combined_SIF <- as.data.frame(rbind(combined_SIF,ligrec_df))

res_mofa_combined <- list(combined_SIF, combined_ATT)
save(res_mofa_combined, file = "results/cosmos/res_mofamoon_combined.RData")
write_csv(combined_SIF, file = "results/cosmos/SIF_mofamoon_combined.csv")
```

To filter the final network based on node properties, the strategy shown here can be used to find interesting ligand-receptor interactions. The final result is saved and later inspected ("Network visualization").

```{r Filter network}
ligands <- combined_ATT[combined_ATT$NodeType == 2,"Nodes"] # 2 for ligand
receptors <- combined_ATT[combined_ATT$NodeType == 3,"Nodes"] # 3 for receptors

combined_SIF_reduced <- combined_SIF[!(combined_SIF$Node2 %in% receptors & !(combined_SIF$Node2 %in% combined_SIF$Node1)),] #no node that is a receptor but not a source
combined_SIF_reduced <- combined_SIF_reduced[!(combined_SIF_reduced$Node2 %in% ligands & !(combined_SIF_reduced$Node2 %in% combined_SIF_reduced$Node1)),] #no node that is a ligand but not a source
combined_SIF_reduced <- combined_SIF_reduced[!(combined_SIF_reduced$Node1 %in% receptors & !(combined_SIF_reduced$Node1 %in% combined_SIF_reduced$Node2)),] #no node that is a receptor but not a regulated target
combined_SIF_reduced <- combined_SIF_reduced[!(combined_SIF_reduced$Node1 %in% ligands & !(combined_SIF_reduced$Node1 %in% combined_SIF_reduced$Node2)),] #no node that is a ligand but not a regulated target 

combined_ATT_reduced <- combined_ATT[combined_ATT$Nodes %in% combined_SIF_reduced$Node1 | combined_ATT$Nodes %in% combined_SIF_reduced$Node2,]

res_mofa_combined_reduced <- list(combined_SIF_reduced, combined_ATT_reduced)
save(res_mofa_combined_reduced, file = "results/cosmos/res_mofamoon_combined_reduced.RData")
write_csv(combined_SIF_reduced, file = "results/cosmos/SIF_mofamoon_combined_reduced.csv")
write_csv(combined_ATT_reduced, file = "results/cosmos/ATT_mofamoon_combined_reduced.csv")
```

Also the specific sub-network of a gene (e.g. MYC) can be extracted by taking into account n-step-distant nodes (here: 4 steps).

```{r Extract sub-network, eval=FALSE}
combined_SIF_reduced <- read_csv(file = "results/cosmos/SIF_mofamoon_combined_reduced.csv")
combined_ATT_reduced <- read_csv(file = "results/cosmos/ATT_mofamoon_combined_reduced.csv")
sig_prots <- combined_SIF_reduced[,c(1,4,2,3)]
names(sig_prots) <- c("source","interaction","target","sign")
background <- unique(c(sig_prots$source,sig_prots$target))

subnetwork_per_gene <- list()

for(node in background){
  
  SIF <- cosmosR:::keep_controllable_neighbours(sig_prots, n_steps = 4, input_nodes = node) #keeps the nodes in network that are no more then n_steps away from the starting nodes
  ATT <- combined_ATT_reduced[combined_ATT_reduced$Nodes %in% SIF$source | combined_ATT_reduced$Nodes %in% SIF$target,] #keep supplementing information for nodes
  
  write_csv(ATT, file = paste0("results/cosmos/subnetwork/ATT_subnetwork_",node,".csv"))
  write_csv(SIF, file = paste0("results/cosmos/subnetwork/SIF_subnetwork_",node,".csv"))
  
  subnetwork_per_gene[[node]] <- list("SIF" = SIF, "ATT" = ATT)
  save(subnetwork_per_gene, file = "results/cosmos/subnetwork/subnetwork_per_gene.RData")
}
```

### Pathway enrichment analysis

A potential downstream analysis of the biological network could be to search for over-represented pathways. The tutorial ./Pathway_enrichment_analysis.Rmd explains possible steps.

### Network visualization via CytoScape

Further, we can analyze the biological network by visualization through CytoScape. A tutorial for that is given under ./RCytoScape_tutorial.Rmd. 

Here, a potential output of this analysis highlighting a sub pathway is shown.

![alt text](results/cytoscape/sub_pathway_figure.pdf "Title")

```{r Full COSMOS network image, echo = FALSE, message=FALSE, fig.align='center', fig.cap='COSMOS sub signaling pathway', out.height = "460px", out.width='800px'}
# knitr::include_graphics("results/cytoscape/sub_pathway_figure.pdf")
```

### MOON (meta footprint analysis) as a CANRIVAL alternative

```{r Prepare moon input for receptor to TFs and metabs}
##filter expressed genes from PKN
data("meta_network")

meta_network <- meta_network_cleanup(meta_network)

expressed_genes <- as.data.frame(read_csv("data/RNA/RNA_log2_FPKM_clean.csv"))
expressed_genes <- setNames(rep(1,length(expressed_genes[,1])), expressed_genes$Genes)
meta_network_filtered <- cosmosR:::filter_pkn_expressed_genes(names(expressed_genes), meta_pkn = meta_network)

##format metab inputs
metab_inputs <- as.numeric(scale(weights$metab[,4]))
names(metab_inputs) <- row.names(weights$metab)

metab_to_HMDB <- as.data.frame(
  read_csv("data/metabolomic/MetaboliteToHMDB.csv"))
metab_to_HMDB <- metab_to_HMDB[metab_to_HMDB$common %in% names(metab_inputs),]
metab_inputs <- metab_inputs[metab_to_HMDB$common]
names(metab_inputs) <- metab_to_HMDB$HMDB

metab_inputs <- cosmosR::prepare_metab_inputs(metab_inputs, compartment_codes = c("m","c"))

#prepare upstream inputs
upstream_inputs <- c(rec_inputs) #the upstream input should be filtered for most significant

downstream_inputs <- c(metab_inputs, ligrec_TF_moon_inputs$TF) #the downstream input should be complete

upstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(upstream_inputs, meta_network_filtered)
downstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(downstream_inputs, meta_network_filtered)

#Filter inputs and prune the meta_network to only keep nodes that can be found downstream of the inputs
#The number of step is quite flexible, 7 steps already covers most of the network

n_steps <- 6

# in this step we prune the network to keep only the relevant part between upstream and downstream nodes
meta_network_filtered <- cosmosR:::keep_controllable_neighbours(meta_network_filtered
                                                       , n_steps, 
                                                       names(upstream_inputs_filtered))
downstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(downstream_inputs_filtered, meta_network_filtered)
meta_network_filtered <- cosmosR:::keep_observable_neighbours(meta_network_filtered, n_steps, names(downstream_inputs_filtered))
upstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(upstream_inputs_filtered, meta_network_filtered)

#compress the network to avoid redundant master controllers
meta_network_compressed_list <- compress_same_children(meta_network_filtered, sig_input = upstream_inputs_filtered, metab_input = downstream_inputs_filtered)

meta_network_compressed <- meta_network_compressed_list$compressed_network

node_signatures <- meta_network_compressed_list$node_signatures

duplicated_parents <- meta_network_compressed_list$duplicated_signatures

meta_network_compressed <- meta_network_cleanup(meta_network_compressed)

load(file = "support/dorothea_df.RData")

RNA_input <- as.numeric(weights$RNA[,4])
names(RNA_input) <- gsub("_RNA$","",row.names(weights$RNA))
```

```{r Run moon rec_to_TFmetab}
meta_network_rec_to_TFmetab <- meta_network_compressed

before <- 1
after <- 0
i <- 1
while (before != after & i < 10) {
  before <- length(meta_network_rec_to_TFmetab[,1])
  recursive_decoupleRnival_res <- cosmosR::moon(upstream_input = upstream_inputs_filtered, 
                                                 downstream_input = downstream_inputs_filtered, 
                                                 meta_network = meta_network_rec_to_TFmetab, 
                                                 n_layers = n_steps, 
                                                 statistic = "ulm") 
  
  meta_network_rec_to_TFmetab <- filter_incohrent_TF_target(recursive_decoupleRnival_res, dorothea_df, meta_network_rec_to_TFmetab, RNA_input)
  after <- length(meta_network_rec_to_TFmetab[,1])
  i <- i + 1
}

if(i < 10)
{
  print(paste("Converged after ",paste(i-1," iterations", sep = ""),sep = ""))
} else
{
  print(paste("Interupted after ",paste(i," iterations. Convergence uncertain.", sep = ""),sep = ""))
}

duplicated_parents_df <- data.frame(duplicated_parents)
duplicated_parents_df$source_original <- row.names(duplicated_parents_df)
names(duplicated_parents_df)[1] <- "source"

addons <- data.frame(names(node_signatures)[-which(names(node_signatures) %in% duplicated_parents_df$source_original)]) 
names(addons)[1] <- "source"
addons$source_original <- addons$source

mapping_table <- as.data.frame(rbind(duplicated_parents_df,addons))

recursive_decoupleRnival_res <- merge(recursive_decoupleRnival_res, mapping_table, by = "source")

levels <- recursive_decoupleRnival_res[,c(4,3)]

recursive_decoupleRnival_res <- recursive_decoupleRnival_res[,c(4,2)]
names(recursive_decoupleRnival_res)[1] <- "source"

plot(density(recursive_decoupleRnival_res$score))
abline(v = 1)
abline(v = -1)

solution_network <- reduce_solution_network(decoupleRnival_res = recursive_decoupleRnival_res, 
                                            meta_network = meta_network_filtered,
                                            cutoff = 1.5, 
                                            upstream_input = upstream_inputs_filtered, 
                                            RNA_input = RNA_input, 
                                            n_steps = n_steps)

SIF_rec_to_TFmetab <- solution_network$SIF
names(SIF_rec_to_TFmetab)[3] <- "sign"
ATT_rec_to_TFmetab <- solution_network$ATT

data("HMDB_mapper_vec")

translated_res <- translate_res(SIF_rec_to_TFmetab,ATT_rec_to_TFmetab,HMDB_mapper_vec)

levels_translated <- translate_res(SIF_rec_to_TFmetab,levels,HMDB_mapper_vec)[[2]]

SIF_rec_to_TFmetab <- translated_res[[1]]
ATT_rec_to_TFmetab <- translated_res[[2]]

## Add weights to data
ATT_rec_to_TFmetab <- merge(ATT_rec_to_TFmetab, MOFA_weights, all.x = T)
ATT_rec_to_TFmetab <- merge(ATT_rec_to_TFmetab, feature_weights, all.x = T)
names(ATT_rec_to_TFmetab)[2] <- "AvgAct"

ATT_rec_to_TFmetab$NodeType <- ifelse(ATT_rec_to_TFmetab$Nodes %in% levels_translated[levels_translated$level == 0,1],1,0)

ATT_rec_to_TFmetab$NodeType <- ifelse(ATT_rec_to_TFmetab$Nodes %in% ligrec_df$Node1, 2, ifelse(ATT_rec_to_TFmetab$Nodes %in% ligrec_df$Node2, 3, ifelse(ATT_rec_to_TFmetab$Nodes %in% dorothea_df$source, 4, ATT_rec_to_TFmetab$NodeType))) 

names(SIF_rec_to_TFmetab)[4] <- "Weight"

write_csv(SIF_rec_to_TFmetab,file = "results/cosmos/moon/SIF_rec_TFmetab.csv")
write_csv(ATT_rec_to_TFmetab,file = "results/cosmos/moon/ATT_rec_TFmetab.csv")
```
```{r Prepare moon input for receptor to TFs and metabs}
##filter expressed genes from PKN

expressed_genes <- as.data.frame(read_csv("data/RNA/RNA_log2_FPKM_clean.csv"))
expressed_genes <- setNames(rep(1,length(expressed_genes[,1])), expressed_genes$Genes)
dorothea_PKN_filtered <- cosmosR:::filter_pkn_expressed_genes(names(expressed_genes), meta_pkn = dorothea_PKN)


#prepare upstream inputs

#the upstream input should be filtered for most significant
upstream_inputs <- setNames(TF_weights$feature_weights, TF_weights$Nodes)
upstream_inputs <- upstream_inputs[abs(upstream_inputs) > 2]

lig_inputs <- ligrec_TF_moon_inputs$ligrec
names(lig_inputs) <- gsub("___.+","",names(lig_inputs))
lig_inputs <- tapply(lig_inputs, names(lig_inputs), mean)
ligands <- names(lig_inputs)
lig_inputs <- as.numeric(lig_inputs)
names(lig_inputs) <- ligands
downstream_inputs <- lig_inputs #the downstream input should be complete

upstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(upstream_inputs, dorothea_PKN_filtered)
downstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(downstream_inputs, dorothea_PKN_filtered)

#Filter inputs and prune the meta_network to only keep nodes that can be found downstream of the inputs
#The number of step is quite flexible, 7 steps already covers most of the network

n_steps <- 1

# in this step we prune the network to keep only the relevant part between upstream and downstream nodes
dorothea_PKN_filtered <- cosmosR:::keep_controllable_neighbours(dorothea_PKN_filtered
                                                       , n_steps, 
                                                       names(upstream_inputs_filtered))
downstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(downstream_inputs_filtered, dorothea_PKN_filtered)
dorothea_PKN_filtered <- cosmosR:::keep_observable_neighbours(dorothea_PKN_filtered, n_steps, names(downstream_inputs_filtered))
upstream_inputs_filtered <- cosmosR:::filter_input_nodes_not_in_pkn(upstream_inputs_filtered, dorothea_PKN_filtered)

#compress the network to avoid redundant master controllers
load(file = "support/dorothea_df.RData")

RNA_input <- as.numeric(weights$RNA[,4])
names(RNA_input) <- gsub("_RNA$","",row.names(weights$RNA))
```

```{r run moon TF to lig}
meta_network_TF_lig <- dorothea_PKN_filtered

before <- 1
after <- 0
i <- 1
while (before != after & i < 10) {
  before <- length(meta_network_TF_lig[,1])
  recursive_decoupleRnival_res <- cosmosR::moon(upstream_input = upstream_inputs_filtered, 
                                                 downstream_input = downstream_inputs_filtered, 
                                                 meta_network = meta_network_TF_lig, 
                                                 n_layers = n_steps, 
                                                 statistic = "ulm") 
  
  meta_network_TF_lig <- filter_incohrent_TF_target(recursive_decoupleRnival_res, dorothea_df, meta_network_TF_lig, RNA_input)
  after <- length(meta_network_TF_lig[,1])
  i <- i + 1
}

if(i < 10)
{
  print(paste("Converged after ",paste(i-1," iterations", sep = ""),sep = ""))
} else
{
  print(paste("Interupted after ",paste(i," iterations. Convergence uncertain.", sep = ""),sep = ""))
}

levels <- recursive_decoupleRnival_res[,c(1,3)]

recursive_decoupleRnival_res <- recursive_decoupleRnival_res[,c(1,2)]
names(recursive_decoupleRnival_res)[1] <- "source"

plot(density(recursive_decoupleRnival_res$score))
abline(v = 1)
abline(v = -1)

solution_network <- reduce_solution_network(decoupleRnival_res = recursive_decoupleRnival_res, 
                                            meta_network = as.data.frame(dorothea_PKN_filtered[,c(1,3,2)]),
                                            cutoff = 1.5, 
                                            upstream_input = upstream_inputs_filtered, 
                                            RNA_input = RNA_input, 
                                            n_steps = n_steps)

SIF_TF_lig <- solution_network$SIF
names(SIF_TF_lig)[3] <- "sign"
ATT_TF_lig <- solution_network$ATT

data("HMDB_mapper_vec")

translated_res <- translate_res(SIF_TF_lig,ATT_TF_lig,HMDB_mapper_vec)

levels_translated <- translate_res(SIF_TF_lig,levels,HMDB_mapper_vec)[[2]]

SIF_TF_lig <- translated_res[[1]]
ATT_TF_lig <- translated_res[[2]]

## Add weights to data
ATT_TF_lig <- merge(ATT_TF_lig, MOFA_weights, all.x = T)
ATT_TF_lig <- merge(ATT_TF_lig, feature_weights, all.x = T)
names(ATT_TF_lig)[2] <- "AvgAct"

ATT_TF_lig$NodeType <- ifelse(ATT_TF_lig$Nodes %in% levels_translated[levels_translated$level == 0,1],1,0)

ATT_TF_lig$NodeType <- ifelse(ATT_TF_lig$Nodes %in% ligrec_df$Node1, 2, ifelse(ATT_TF_lig$Nodes %in% ligrec_df$Node2, 3, ifelse(ATT_TF_lig$Nodes %in% dorothea_df$source, 4, ATT_TF_lig$NodeType))) 

names(SIF_TF_lig)[4] <- "Weight"

write_csv(SIF_TF_lig,file = "results/cosmos/moon/SIF_rec_TFmetab.csv")
write_csv(ATT_TF_lig,file = "results/cosmos/moon/ATT_rec_TFmetab.csv")
```

```{r combine moon networks}
combined_SIF_moon <- as.data.frame(rbind(SIF_rec_to_TFmetab, SIF_TF_lig))
combined_SIF_moon <- unique(combined_SIF_moon)

combined_ATT_moon <- as.data.frame(rbind(ATT_rec_to_TFmetab, ATT_TF_lig))
combined_ATT_moon <- combined_ATT_moon %>% group_by(Nodes) %>% summarise_each(funs(mean(., na.rm = TRUE)))
combined_ATT_moon <- as.data.frame(combined_ATT_moon)

ligrec_ressource_addon <- ligrec_ressource[
  ligrec_ressource$source_genesymbol %in% combined_SIF_moon$source &
    ligrec_ressource$target_genesymbol %in% combined_SIF_moon$target
, c(10,12)]
ligrec_ressource_addon$sign <- 1
ligrec_ressource_addon$Weight <- 1
names(ligrec_ressource_addon)[c(1,2)] <- c("source","target")
ligrec_ressource_addon <- unique(ligrec_ressource_addon)

combined_SIF_moon <- as.data.frame(rbind(combined_SIF_moon, ligrec_ressource_addon))

#It appears I may need to only consider direct TF regulations for TF to lig

write_csv(combined_SIF_moon,file = "results/cosmos/moon/combined_SIF_moon.csv")
write_csv(combined_ATT_moon,file = "results/cosmos/moon/combined_ATT_moon.csv")
```
### Further analyses

Further MOFA-COSMOS downstream analyses focusing on analyzing cell line specific MOFA transcriptomics are given in ./Further_MOFA_COSMOS_analyses.Rmd.

```{r Select sub pathway nodes}
## Select nodes 
interesting_nodes <- data.frame("name" = c("VEGFA","HIF1A","PRKCA","BNIP3","DAG1","ACO1","JAK1","ABL1","ITGB1","STAT1","PRKACA","MYC"))
nodes_cellline_weight_filtered <- nodes_cellline_weight[rownames(nodes_cellline_weight) %in% interesting_nodes$name,]
```


### Session info

```{r Session info}
sessionInfo()
```
